### 2016-10-11 lists_conversion_experience.rmd
starting basket similarity

Presentation presents another presentation I saw recently about online grocery shopping at Tesco, a large supermarket chain based in the UK. The lists in this setting are recommendations for further purchases under the title of 'Have you forgotten?'. When you shop for groceries online, the 'Have you forgotten' recommendations appear when at the checkout stage. While the title 'have you forgotten' implies that you will be shown items that could have been or are usually on a shopping list. The working assumption is a temporal one -- the list is incomplete either because you have forgotten to list something, or you didn't know what should be on your list, perhaps because you have never even thought of putting that thing on your list.

@@

But a recommendation system finds it much harder to suggest items that you have not forgotten but may have never bought. If you do buy recommended items, then the recommender system has been effective. It has led to a **'conversion.'** From a supermarket perspective, the conversion should be high-value. There is little value in recommending a whole lot of small and cheap items such as rubber bands or toothpicks. So it might need to choose amongst different recommendations and weight high value items differently to low value items. In either case, 'conversion'  is the valued event.

If we are interested in lists from various angles -- as asignifying semiotics as Maurizio Lazzarato calls it in his _Signs and Machines_ [@Lazzarato_2014] or as elements in 'a new order of spatio-temporal continuity for forms of economic, political and cultural life' as Celia Lury terms in her account of topological turn [@TBA], or in terms of the predictive conditions of contemporary experience [@Hansen_2015a] -- then contemporary list practices such as recommendations figure as sites of intersection between different forms.

@@

Much of my paper, and my responses to the presentation I saw from a data scientist  will focus on how the ordering that is done in a recommendation lists relates to the problem of social order more generally.  The main narrative of the presentation concerned a shift from a set-based model for making lists to a predictive, probabilistic way of making lists in almost realtime and dealing with lists on a scale that is often difficult to grasp concretely. The change in list-making associated with learned models is increasingly widespread, and

I respond to the presentation in two ways -- firstly by recounting, contextualising and commenting on some of the main steps in making the 'Have your forgotten?' list  as a example of the sense-making done by data scientists and customer analytics teams in many places; and secondly, by trying to step into the list-making practices  as an occasion to intervene and participate in the ontological conversion practices they engage in. This is no mere methodological problem if, as Mark Hansen has recently suggested,

@@

## 1. The context of the presentation - Consumption and shopping

@@

Shreena described her work at DunnHumby in terms of a shift in models from a 'rules-based list' to a 'relevancy model.' 
This was a significant change in how the 'Have You Forgotten' list was constructed from one kind of model to another. For me, the shift in models matters since it amounts to a much more probabilistic structuring of lists.
I assume that the rules-based model derived from the early 1990s work on the `apriori` algorithm by Rakesh Agrawal and Ramakrishnan Srikanti at IBM Research Alameda [@Agrawal_1994].
The `apriori` algorithm finds sets of items that commonly occur in transactions. 
These commonly occurring sets are expressed as association rules. 
For instance, using the `Groceries` dataset in the `R` package `arules,` we can readily see how Walmart might have discovered  -- but didn't -- the fabled association between diapers and beer.  [http://www.dssresources.com/newsletters/66.php]
It is a key algorithm in 'market basket analysis' and attempts to re-structure shopping lists.

@@

The association rule-based model for the `Groceries` dataset suggests some of the characteristic ordering practices of data-mining approaches to lists. 
The association between items is expressed in the form of _rules_. These rules, a term derived from decision support literature, define sets of associations between items. 
As a set theory-based practice, the association rules define likely sets of items or shopping baskets in terms of decreasing probability.   
`r inspect(rules[1])` is a stronger association than `r inspect(rules[2])`.
So any recommendation in the 'Have You Forgotten' list should probably put them in that order. 
Association rules and the `apriori` algorithm present social order in terms of sets of common associations. 
This means that that the algorithm can be understood in terms of the set-based multiples proposed by Alain Badiou.

@@

Badiou in his account of multiples highlight the contrast between two different ways of grasping a multiple -- in terms of the 'numerical one' and the 'vital One' [@Badiou_2004, 72], between the 'closed' and the 'open' respectively.  
The `apriori` algorithm, might be seen as an attempt to deal with the excess of things in a Tesco supermarket, and the lack of interest or engagement that people have with them by identifying subsets whose membership is stronger by virtue of the other sets of associations that people bring with them.   
Association is stronger when it derives from an excess of different ways of being in sets.

@@

Shreena observed that the 'Have you forgotten' lists matter to people, and we can see why in the light of this proliferation of things in the list. A `r total` list of items is quite hard to list let alone read. 
Shreena said that some people log on and go straight to the checkout, allowing the recommendation system to supply them with the suggestions that they then move to the shopping basket.
This is the online equivalent of using an old shopping list and only adding new items when needed. 
It is also a strategy to avoid having to view too many lists, since the only way Tesco can show people what they might buy is by showing them lists of things (e.g. chicken products), but these lists are often too long to work with. For instance, Tesco has more than 1000 chicken products on its site.   
People inhabit lists and work with them no matter how they are made; thus vindicating ethnomethodologists assertion that social order is the product of ongoing sense-making practices. 
But it also suggests a transformation in the way that lists exist in the world.

@@

Shreena described the data they on grocery purchases. 'We have 52 weeks of purchase data for each customer.' 
We know that the construction of lists has been transformed by recommendation systems.
These transformations have been widely discussed in the context of big data, where questions of who has how much data are usually most prominent.
In terms of the ways that lists order the world, the amount of data is not so amount as the way in which things in the world take on a list-like existence. 
I want to dwell on this idea for a while. We have seen two different ways in which lists might be constructed in a recommender system. 
The first uses history and existence of previous lists, including the customer's own list, to populate a new list. 
The second uses many shopping lists to identify associations between things and then uses the strongest associations to recommend additions to a current list. 
The 'relevancy model,' however, tries to do something different. 
The problem that the proliferation of things on the shelves of supermarket or grocery warehouse confronts is partly combinatorial.
With  `r total` products (actually Shreena mentions 200,000 products) can be combined in truly astronomical numbers of combinations.
If a typical shopping list has 20 items, then there are `r format(choose(total, 20), 2)` possible combinations.
Some of this vast number of possible shopping lists is highly unlikely in terms of shopping lists and histories of shopping, but potentially highly interesting and value as recommendation.

@@

In some ways, Hansen's account resembles the Badiouean ontology of the multiple in which a closed set is surrounded and conditioned by an open set that is full of many more potentials. However, if we think of the  new model that Shreena describes as the delimited (or delimiting?) system within   the larger surplus of sensibility, how that help us understand what the 'Have you forgotten?' recommendations do?

@@

## 6. Gain: use good features to build models 
- what counts as a good feature? See Domingos on this
- move from Dunnhumby's ClubCard focus to much more general world of predictive models (e.g. Facebook use a million models, etc.)
- the model counts for surprisingly little -- at least, they all seem similar ... rather what counts is all that stuff on working with substitutions, similiarities and complementarities for each person.

It was surprising that Shreena spoke relatively briefly about the new models. She showed a graph that compared the old rule-based model with some of the new alternatives -- logistic regression, random forests, and a few others. She dismissed most of the models quite quickly and focused only on one, the logistic regression model. Or more accurately, she just said that logistic regression did better than other, and there was little more to discusson on that front.  The measure of doing better here is called precision, and it measures the proportion of the recommended products that are actually purchased.

@@

@@

Predictive models, we might say, encounter their own surplus of sensibility in the data as they search for the most accurate predictions. A  very brief comment from Shreena -- 'we have lots of zeros' -- points to the problem here. Each of the millions of Tesco online grocery shoppers has bought some selection of the 200,000 products. But the product-customer matrix, the basic vector-space in which all recommendation systems operate, remain very sparse and unpopulated.

@@

So the raw data -- the 52 weeks of purchases for each customer -- can be aggregated to increase the density of the data in ways that helps the model. Shreena mentions some of the 'features' used in the new model but somewhat strangely at this point, her presentation turns toward the difficulties of measuring the success of the model and the infrastructural limitations on making recommendations in ways that are more alive to what is happening right now.

@@

The discussion of imperfection was brief - it incorporated the notion of recommendations weighted by price and the notion of precision (or the proportion of products actually purchased.). Shreena simply said that this metric was imperfect, but without elaborating why. Instead she moved to describe some of the ways in which the recommendation system worked in practice.

@@

While much discussion of recent media has highlighted its realtime operation, Shreena's description of the infrastructure that implements the new model suggested a different temporal patterning. While raw data from Tesco Online feeds into the model every hour, it seems that recommendation list for each customer is only generated once a week. This makes sense. Customers only shop online every few days at the most, and in some cases, only every few weeks. To continually update the top 200 recommendations for every customer would be computationally expensive.

Shreena did not say much else about the infrastructure but the 200 recommendations generated each week are certainly never seen by the customers. The realtime engine that actually generates results shown on the Tesco Online website only shows 5 after 'applying business rules.' Again, this might be seen as corroborating Hansen and Badiou's insistence on the necessarily limited grasp of any prediction system on the situation it encounters. Even the new model, situated amidst its experimental setup cannot fully encompass the surplus of sensibility it encounters.

@@

## 12. Similarity, substitution and complementarity

- interesting contrasts here between the relations of things in a list
- whereas the `arule` approach compares lists looking for similiarities and ways of completing lists, the more recent approaches look for micrological propensities in the list that relate to a particular time and place, amidst a continuously differentiable pathways between things ... 
- how do things associate with each other in complementarities, substitutions, etc.

>>>>

Presentation presents another presentation I saw recently about online grocery shopping at Tesco, a large supermarket chain based in the UK. The lists in this setting are recommendations for further grocery purchases under the title of 'Have you forgotten?'. When you shop for groceries online, the 'Have you forgotten' recommendations appear  at the checkout stage. While the title 'have you forgotten' implies that you will be shown items that could have been or are usually on a shopping list, the working assumption is a temporal one -- the list is incomplete either because you have forgotten to list something, or you didn't know what should be on your list, perhaps because you have never even thought of putting that thing on your list.

@@

But a recommendation system finds it much harder to suggest items that you have not forgotten but may have never bought. If you do buy recommended items, then the recommender system has been effective in shunting the situation differently. It has led to a **'conversion.'** From a supermarket perspective, the conversion should be high-value. There is little value in recommending a whole lot of small and cheap items such as rubber bands or toothpicks. So it might need to choose amongst different recommendations and weight high value items differently to low value items. In either case, 'conversion'  is the valued event.

We could regard lists from various angles -- as asignifying semiotics as Maurizio Lazzarato calls it in his _Signs and Machines_ [@Lazzarato_2014] or as elements in 'a new order of spatio-temporal continuity for forms of economic, political and cultural life' as Celia Lury terms in her account of topological turn [@TBA]. In this paper, I focus on lists and their propensities in terms of the predictive conditions of contemporary experience [@Hansen_2015a].  List practices such as recommendations figure as sites of intersection between people, infrastructures, things, places and predictions.

@@

Much of my paper, and my responses to the presentation I saw from a data scientist  will focus on how the ordering that is done in a recommendation lists relates to the problem of social order more generally.  The main narrative of the presentation concerned a shift from a set-based model for making lists to a predictive, probabilistic way of making lists in almost realtime and dealing with lists on a scale that is often difficult to grasp concretely. The change in list-making associated with learned models is increasingly widespread, and relies on a probabilistic treatment of people and things.

I respond to the presentation in two ways. First, by recounting, contextualising and commenting on some of the main steps in making the 'Have your forgotten?' list  as a example of the sense-making done by data scientists and customer analytics teams in many places. Many of these steps were mentioned only briefly in the presentation, and they are largely taken for granted as part of predictive analytic practice.  Second, by trying to step into the list-making practices  as an occasion to intervene and participate in the ontological conversion practices they engage in. This is no mere methodological problem if, as Mark Hansen has recently suggested,

@@

## 1. The context of the presentation: 'statistics and operations research in industry'

@@

Patel described her work at DunnHumby in terms of a shift in models from a 'rules-based list' to a 'relevancy model.' This was a significant change in how the 'Have You Forgotten' list was constructed from one kind of model to another. For me, the shift in models matters since it amounts to a much more probabilistic structuring of lists.

I assume that the rules-based model derived from the early 1990s work on the `apriori` algorithm by Rakesh Agrawal and Ramakrishnan Srikanti at IBM Research Alameda [@Agrawal_1994]. The `apriori` algorithm finds sets of items that commonly occur in transactions. These commonly occurring sets are expressed as association rules. For instance, using the `Groceries` dataset in the `R` package `arules,` we can readily see how Walmart might have discovered  -- but didn't -- the fabled association between diapers and beer.  [http://www.dssresources.com/newsletters/66.php]
The `apriori` algorithm is widely used in 'market basket analysis' and attempts to re-structure shopping lists in terms of associations between items that are frequently bought together.

@@

The association rule-based model for the `Groceries` dataset suggests some of the characteristic ordering practices of data-mining approaches to lists. 
The association between items is expressed in the form of _rules_. These rules, a term derived from decision support literature, define sets of associations between items. As a set theory-based practice, the association rules define likely sets of items or shopping baskets in terms of decreasing probability.   `r inspect(rules[1])` is a stronger association than `r inspect(rules[2])`. So any recommendation in the 'Have You Forgotten' list should probably put them in that order.

Association rules and the `apriori` algorithm present social order in terms of sets of common associations. This means that that the algorithm can be understood in terms of the set-based multiples proposed by Alain Badiou:

@@

Badiou in his account of multiples highlight the contrast between two different ways of grasping a multiple -- in terms of the 'numerical one' and the 'vital One' [@Badiou_2004, 72], between the 'closed' and the 'open' respectively.

The `apriori` algorithm, might be seen as an attempt to deal with the excess of things in a Tesco supermarket, and the lack of interest or engagement that people have with them by identifying subsets whose membership is stronger by virtue of the other sets of associations that people bring with them.   
Association is stronger when it derives from an excess of different ways of being in sets.

@@

Patel observed that the 'Have you forgotten' lists matter to people, and we can see why in the light of this proliferation of things in the list. A `r total` list of items is quite hard to list let alone read. 
Patel said that some people log on and go straight to the checkout, allowing the recommendation system to supply them with the suggestions that they then move to the shopping basket.
This is the online equivalent of using an old shopping list and only adding new items when needed. 
It is also a strategy to avoid having to view too many lists, since the only way Tesco can show people what they might buy is by showing them lists of things (e.g. chicken products), but these lists are often too long to work with. For instance, Tesco has more than 1000 chicken products on its site.   
People inhabit lists and work with them no matter how they are made; thus vindicating ethnomethodologists assertion that social order is the product of ongoing sense-making practices. 
But it also suggests a transformation in the way that lists exist in the world. They are less the product of a discrete agent than an effect of an environment teeming with possible combinations and complementaries, a surplus of possible purchases that the supermarket needs to continually maintain but also filter in order to make it possible for something to happen.

@@

Patel described the data they on grocery purchases. 'We have 52 weeks of purchase data for each customer.' 
We know that the construction of lists has been transformed by recommendation systems.
These transformations have been widely discussed in the context of big data, where questions of who has how much data are usually most prominent.
In terms of the ways that lists order the world, the amount of data is not so amount as the way in which things in the world take on a list-like existence. 
I want to dwell on this idea for a while. We have seen two different ways in which lists might be constructed in a recommender system. 
The first uses history and existence of previous lists, including the customer's own list, to populate a new list. 
The second uses many shopping lists to identify associations between things and then uses the strongest associations to recommend additions to a current list. 
The 'relevancy model,' however, tries to do something different. 
The problem that the proliferation of things on the shelves of supermarket or grocery warehouse confronts is partly combinatorial.
With  `r total` products (actually Patel mentions 200,000 products) can be combined in truly astronomical numbers of combinations.
If a typical shopping list has 20 items, then there are `r format(choose(total, 20), 2)` possible combinations.
Some of this vast number of possible shopping lists is highly unlikely in terms of shopping lists and histories of shopping, but potentially highly interesting and value as recommendation.

@@

In some ways, Hansen's account resembles the Badiouean ontology of the multiple in which a closed set is surrounded and conditioned by an open set that is full of many more potentials. However, if we think of the  new model that Patel describes as the delimited (or delimiting?) system within   the larger surplus of sensibility, how that help us understand what the 'Have you forgotten?' recommendations do?

@@

## 6. Gain: use good features to build models

- what counts as a good feature? See Domingos on this
- move from Dunnhumby's ClubCard focus to much more general world of predictive models (e.g. Facebook use a million models, etc.)
- the model counts for surprisingly little -- at least, they all seem similar ... rather what counts is all that stuff on working with substitutions, similiarities and complementarities for each person.

It was surprising that Patel spoke relatively briefly about the new models. She showed a graph that compared the old rule-based model with some of the new alternatives -- logistic regression, random forests, and a few others. She dismissed most of the models quite quickly and focused only on one, the logistic regression model. Or more accurately, she just said that logistic regression did better than other, and there was little more to discuss on that front.  The measure of doing better here is called precision, and it measures the proportion of the recommended products that are actually purchased.

@@

The fact that so little was said about the models themselves suggests that they are now largely taken for granted in recommendation systems. The models she mentioned are all classifiers, predictive models that treat a set of variables in a datasets as a vector space in which good classification means finding a way of defining a decision boundary between points that belong to different classes of things. Here the classes of things are binary: `recommended` or `not recommended.` Some classifier models, including logistic regression, generate a probability rather than simply a classification.

@@

Predictive models, we might say, encounter their own surplus of sensibility in the data as they search for the most accurate predictions. A  very brief comment from Patel -- 'we have lots of zeros' -- points to the problem here. Each of the millions of Tesco online grocery shoppers has bought some selection of the 200,000 products. But the product-customer matrix, the basic vector-space in which all recommendation systems operate, remain very sparse and unpopulated.

@@

So the raw data -- the 52 weeks of purchases for each customer -- can be aggregated to increase the density of the data in ways that helps the model. Patel mentions some of the 'features' used in the new model but somewhat strangely at this point, her presentation turns toward the difficulties of measuring the success of the model and the infrastructural limitations on making recommendations in ways that are more alive to what is happening right now.

@@

The discussion of imperfection was brief - it incorporated the notion of recommendations weighted by price and the notion of precision (or the proportion of products actually purchased.). Patel simply said that this metric was imperfect, but without elaborating why. Instead she moved to describe some of the ways in which the recommendation system worked in practice.

@@

While much discussion of recent media has highlighted its realtime operation, Patel's description of the infrastructure that implements the new model suggested a different temporal patterning. While raw data from Tesco Online feeds into the model every hour, it seems that recommendation list for each customer is only generated once a week. This makes sense. Customers only shop online every few days at the most, and in some cases, only every few weeks. To continually update the top 200 recommendations for every customer would be computationally expensive.

Patel did not say much else about the infrastructure but the 200 recommendations generated each week are certainly never seen by the customers. The realtime engine that actually generates results shown on the Tesco Online website only shows 5 after 'applying business rules.' Again, this might be seen as corroborating Hansen and Badiou's insistence on the necessarily limited grasp of any prediction system on the situation it encounters. Even the new model, situated amidst its experimental setup cannot fully encompass the surplus of sensibility it encounters.

@@

## 12. Similarity, substitution, complementarity and micrological propensities

![Basket similarity](figure/patel_basket_similarity.png)

What does it matter that recommendations are relevant? Relevance is closely connected to propensities in the predictive condition. From DunnHumby's standpoint, the relevance of a recommendation relates directly to conversion. From the standpoint of the ontological conversion, and  if Hansen is right, the real or actual fabric of prediction resides in the propensities of things, what he calls the 'complex and diffuse calculus' that only subsequently and provisionally congeals as a conversion event. The 'micrological propensities' of things feed directly into an sense of the relevance of recommendations.

How does the DunnHumby model deal with the micrological propensities of things? In her presentation, Patel  spent quite a lot of time addressing the problem of basket similarity in terms of substitutions and complementarities between things. She concludes by saying 'basket similarity is the future' and 'complementarities drive conversion.'

Whereas the association rules focus on complementarities -- if you buy bread, you might want milk, if you buy beer, you might want diapers -- the substitutions look for micrological propensities that relate to gradations of preference, and variations in time and place.  Micrological propensities in the list relate to a particular time and place, amidst a continuously differentiable pathways between things.

How do things associate with each other in  substitutions and complementarities? Patel's comments on this are very brief. Like the mention of `hadoop`, they imply quite extensive but relatively well-established data practices. She speaks for instance of a 'design matrix of 14,000 columns' and the use of `L1 regularization` to drive coefficients down to zero.

```{r biscuits, echo=FALSE, cache=TRUE}
    library(dplyr)
    library(ggplot2)
    library(readr)
    bisc = read_tsv('data/biscuits.csv')
    bisc %>% select(price, name)  %>% order(price, decreasing=TRUE)    
    bisc[order(bisc$price, decreasing=TRUE), c('name','price')]
```
We can imagine this in terms of biscuits.
